---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# scorcher

<!-- badges: start -->
<!-- badges: end -->

`scorcher` provides a high-level, declarative workflow for building R `torch` 
models. A `scorcher` model is written as a graph: declare the inputs, add 
layers or functions in the order data flows through them, and mark the output 
nodes. That graph remains inspectable throughout modeling, training, plotting, 
and auditing.

In `scorcher`, a graph is a diagram-like record of the model. Each named step 
in the graph is a node, and the connections between nodes show how information
moves from inputs to predictions.

Models are defined with `scorcher` graph functions such as `scorch_layer()`,
`scorch_function()`, `scorch_concat()`, and `scorch_add_skip()`. The same 
graph can be trained with the lightweight `fit_scorch()` loop or converted 
with `as_torch()` for use with `luz`.

## Why Neural Networks Fit This Workflow

Neural networks are flexible prediction models built from many small
mathematical steps. Each step transforms the data a little: one layer might
combine input variables, another might add a nonlinear bend, another might
drop information during training to reduce overfitting, and a final layer
turns the learned representation into a prediction.

That step-by-step structure makes neural networks a natural fit for tidy,
pipe-based R code. With Scorcher, a model reads from top to bottom in the same
order the data moves through it:

- `initiate_scorch()` starts the model object.
- `scorch_input()` declares what data enters the model.
- `scorch_layer()`, `scorch_function()`, `scorch_block()` add modeling steps.
- `scorch_output()` declares what the model should return.
- `compile_scorch()` and `fit_scorch()` train the model.

This keeps the code close to the mental model: start with data, pass it 
through named steps, inspect the architecture, train, and evaluate. For 
applied research workflows, that makes the model easier to read, easier 
to plot, and easier to audit later.

## Installation

```{r install}
# install.packages("pak")
pak::pak("jtleek/scorcher")
```

## A Small Model Graph

A Scorcher model starts with a dataloader: an object that feeds small groups 
of examples to the model during training. The graph below takes four numeric
features, passes them through a hidden layer and a nonlinear step, applies
dropout, and ends with a one-number prediction. Dropout randomly hides some
intermediate values during training, which can help the model rely less on 
any single feature pattern.

```{r example}
library(torch)
library(scorcher)

x <- torch_randn(100, 4)
y <- torch_randn(100, 1)
dl <- scorch_create_dataloader(x, y, batch_size = 16)

model <- initiate_scorch(dl) |>
  scorch_input(features) |>
  scorch_layer(linear, in_features = 4, out_features = 16) |>
  scorch_layer(relu) |>
  scorch_dropout(p = 0.1) |>
  scorch_layer(linear, in_features = 16, out_features = 1, .name = prediction) |>
  scorch_output()

scorch_spec(model)
validate_scorch_graph(model)
plot(model, detail = "full", input_shapes = list(features = "batch x 4"))
plot(model, detail = "simple")
```

Unnamed nodes are deterministic: `linear_1`, `relu_1`, `dropout_1`, and so on.
Use `.name = prediction` when a stable semantic name matters. Use `.from = ...`
to branch or merge graph paths.

## Graph Blocks

Blocks are useful when several layers form one conceptual unit. In this 
example, `scorch_block()` builds a small residual MLP block from ordinary 
torch layer types. The input to the block is added back to the block output 
because `residual = TRUE`. A residual connection is a shortcut: the block 
learns a change to apply while still keeping the original incoming information.

```{r block}
model <- initiate_scorch(dl) |>
  scorch_input(features) |>
  scorch_layer(linear, in_features = 4, out_features = 16, .name = hidden) |>
  scorch_block(
    linear(in_features = 16, out_features = 16),
    gelu(),
    residual = TRUE,
    .from = hidden,
    .name = residual
  ) |>
  scorch_layer(linear, in_features = 16, out_features = 1, .name = prediction) |>
  scorch_output(prediction)
```

## Training And Reproducibility

After the graph structure is clear, `compile_scorch()` attaches the loss and
optimizer, and `fit_scorch()` trains the model. The loss is the number the
model tries to make smaller, and the optimizer is the rule used to update model
weights. The fitted object records the training history, device, seed, and
runtime metadata.

```{r train}
scorch_fit <- model |>
  compile_scorch(
    loss_fn = nn_mse_loss(),
    optimizer_fn = optim_adam,
    optimizer_params = list(lr = 1e-3)
  ) |>
  fit_scorch(num_epochs = 5, seed = 1)

run <- scorch_snapshot(scorch_fit, data = x)
scorch_audit(run)
autoplot(run, type = "history")
```

The same graph can also be converted for use with `luz`. Use a regular torch
dataloader that yields inputs and the answers the model should learn to
predict.

```{r luz}
torch_module <- as_torch(model)
luz_dl <- torch::dataloader(
  torch::tensor_dataset(x, y),
  batch_size = 16
)

luz_fit <- torch_module |>
  luz::setup(
    loss = nn_mse_loss(),
    optimizer = optim_adam
  ) |>
  luz::fit(data = luz_dl, epochs = 5)
```

## Advanced Examples

See:

- `vignette("scorch_layer")` for lower-level graph construction examples.
- `vignette("palmer-penguins")` for a complete tabular classification
  workflow with preprocessing, fitting, auditing, and evaluation.
- `vignette("mnist")` for an image-classification workflow with convolutional
  layers, which are layers designed to find local image patterns.
- `vignette("diffusion")` for a `scorcher` reproduction of
  [tanelp/tiny-diffusion](https://github.com/tanelp/tiny-diffusion).
- `vignette("mingpt")` for a `scorcher` mirror of minGPT's
  [chargpt tiny-shakespeare example](https://github.com/karpathy/minGPT/blob/master/projects/chargpt/readme.md).

