---
title: "A 2D Diffusion Workflow with Scorcher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A 2D Diffusion Workflow with Scorcher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette reproduces the core workflow from
[tanelp/tiny-diffusion](https://github.com/tanelp/tiny-diffusion) using
`scorcher` graph functions. The original project is a small diffusion model for
2D datasets. A diffusion model learns to reverse a gradual noising process:
training adds random noise to points, and the model learns how much noise was
added so it can remove that noise during generation.

```{r setup}
library(torch)
library(scorcher)
library(datasauRus)
library(tidyverse)
```

## Data

The default tiny-diffusion experiment trains on the Datasaurus Dozen `dino`
dataset. The scaling below follows that setup.

```{r data}
set.seed(123)
nobs <- 8000
dino <- datasauRus::datasaurus_dozen
dino <- dino[dino$dataset == "dino", c("x", "y")]
dino <- dino[sample(seq_len(nrow(dino)), nobs, replace = TRUE), ]
dino$x <- ((dino$x + rnorm(nobs, sd = 0.15)) / 54 - 1) * 4
dino$y <- ((dino$y + rnorm(nobs, sd = 0.15)) / 48 - 1) * 4
x <- as.matrix(dino)

input <- output <- torch_tensor(x, dtype = torch_float())
dl <- scorch_create_dataloader(input, output, batch_size = 32)
```

The dataloader feeds the model small groups of points during training.

## Diffusion Feature Nodes

The tiny-diffusion model turns `x`, `y`, and the diffusion timestep `t` into
learned numeric features, joins those features together, and passes them
through a small residual MLP. An MLP is a standard fully connected neural
network. A residual block has a shortcut connection that keeps the incoming
information while learning a change to apply.

```{r block}
input_column <- function(x, column) x[, column]
emb_size <- 128
hidden_size <- 128
```

## Model Graph

The timestep is declared with `scorch_input(timesteps)`, just like any other
model input. The graph can be inspected before training so the model structure
is visible.

```{r model}
model <- initiate_scorch(dl) |>
  scorch_input(input) |>
  scorch_input(timesteps) |>
  scorch_function(input_column, column = 1, .from = input, .name = input_x) |>
  scorch_function(input_column, column = 2, .from = input, .name = input_y) |>
  scorch_layer(PositionalEmbedding, size = emb_size, type = "sinusoidal",
               scale = 25, .from = input_x, .name = input_x_emb) |>
  scorch_layer(PositionalEmbedding, size = emb_size, type = "sinusoidal",
               scale = 25, .from = input_y, .name = input_y_emb) |>
  scorch_layer(PositionalEmbedding, size = emb_size, type = "sinusoidal",
               .from = timesteps, .name = timestep_emb) |>
  scorch_concat(.from = c(input_x_emb, input_y_emb, timestep_emb),
                dim = 2, .name = features) |>
  scorch_layer(linear, in_features = emb_size * 3, out_features = hidden_size,
               .from = features, .name = input_projection) |>
  scorch_layer(gelu, .name = input_activation) |>
  scorch_block(
    linear(in_features = hidden_size, out_features = hidden_size),
    gelu(),
    residual = TRUE,
    repeats = 3,
    .name = residual_mlp,
    .from = input_activation
  ) |>
  scorch_layer(linear, in_features = hidden_size, out_features = 2,
               .from = residual_mlp,
               .name = predicted_noise) |>
  scorch_output(predicted_noise)

scorch_spec(model)
plot(
  model,
  input_shapes = list(input = "batch x 2", timesteps = "batch")
)
```

## Training

`scorch_2d_diffusion_train()` prepares each training batch by adding random
noise to the points. It returns the noisy points, the timesteps, and the noise
the model should learn to predict.

```{r train}
noise_scheduler <- NoiseScheduler(num_timesteps = 50L, beta_schedule = "linear")

scorch_fit <- model |>
  compile_scorch(
    loss_fn = nn_mse_loss(),
    optimizer_fn = optim_adamw,
    optimizer_params = list(lr = 1e-3)
  ) |>
  fit_scorch(
    num_epochs = 200,
    preprocess_fn = scorch_2d_diffusion_train,
    noise_scheduler = noise_scheduler,
    seed = 1,
    verbose = TRUE
  )

run <- scorch_snapshot(scorch_fit, data = x)
scorch_audit(run)
autoplot(run, type = "history")
```

For diffusion models, `fit_scorch()` is convenient because
`scorch_2d_diffusion_train()` samples a timestep, adds noise, and creates the
noise target for each training batch. A `luz` workflow can use
`as_torch(model)` with a dataloader that already yields the same pieces: noisy
coordinates, timesteps, and the noise to predict.

## Generate Dinosaur Samples from Noise

After training, generation starts from random 2D Gaussian noise, meaning random
points drawn from a bell-shaped distribution. At each reverse diffusion step,
the model predicts the noise in the current points; the scheduler then uses
that prediction to move one step closer to the learned dinosaur-shaped point
cloud.

```{r sample}

set.seed(1)

n_samples <- 2000L

generated <- torch_randn(n_samples, 2)

for (t in rev(seq_len(noise_scheduler$num_timesteps) - 1L)) {
  t_tensor <- torch_tensor(rep(t, n_samples), dtype = torch_long())
  predicted_noise <- scorch_predict(
    scorch_fit,
    input = list(input = generated, timesteps = t_tensor)
  )
  generated <- noise_scheduler$step(predicted_noise, t, generated)
}

generated_xy <- as.data.frame(as_array(generated))

names(generated_xy) <- c("x", "y")

ggplot(generated_xy, aes(x = x, y = y)) +
  geom_point(size = 0.35, alpha = 0.65) +
  coord_equal(xlim = c(-6, 6), ylim = c(-6, 6)) +
  theme_void()
```

With enough epochs, these samples should recover the outline of the Datasaurus
`dino` point cloud from pure noise.
