---
title: "A Tiny GPT-Style Model with Scorcher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Tiny GPT-Style Model with Scorcher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette mirrors the character-level GPT example from
[karpathy/minGPT](https://github.com/karpathy/minGPT) using `scorcher` graph
functions. In minGPT, the `projects/chargpt` example trains a character-level
language model from a user-provided `input.txt`, and the README points to
tiny-shakespeare as the example text file. The model path is the same GPT
pattern: turn characters and positions into numeric representations, let the
model attend to earlier characters in the sequence, and predict the next
character at each position.

```{r setup}
library(torch)
library(scorcher)
```

## Tiny Shakespeare Character Data

The model works at the character level. Each unique character in the text gets
an integer id, and each training example contains a fixed-length context plus
the next character at every position in that context.

```{r data}
input_path <- "input.txt"

if (!file.exists(input_path)) {
  download.file(
    "https://raw.githubusercontent.com/karpathy/char-rnn/master/data/tinyshakespeare/input.txt",
    input_path
  )
}

text <- paste(readLines(input_path, warn = FALSE), collapse = "\n")
chars <- sort(unique(strsplit(text, "")[[1]]))
stoi <- setNames(seq_along(chars), chars)
encoded <- unname(stoi[strsplit(text, "")[[1]]])
split_at <- floor(0.9 * length(encoded))
train_encoded <- encoded[seq_len(split_at)]
validation_encoded <- encoded[(split_at + 1L):length(encoded)]

block_size <- 128L
make_xy <- function(encoded, block_size) {
  starts <- seq_len(length(encoded) - block_size)
  x <- do.call(rbind, lapply(starts, function(i) encoded[i:(i + block_size - 1)]))
  y <- do.call(rbind, lapply(starts, function(i) encoded[(i + 1):(i + block_size)]))
  list(
    x = torch_tensor(x, dtype = torch_long()),
    y = torch_tensor(y, dtype = torch_long())
  )
}

xy <- make_xy(train_encoded, block_size)
validation_xy <- make_xy(validation_encoded, block_size)
dl <- scorch_create_dataloader(
  xy$x,
  xy$y,
  batch_size = 64,
  shuffle = TRUE,
  input_dtype = torch_long(),
  output_dtype = torch_long()
)
validation_dl <- scorch_create_dataloader(
  validation_xy$x,
  validation_xy$y,
  batch_size = 64,
  shuffle = FALSE,
  input_dtype = torch_long(),
  output_dtype = torch_long()
)
```

## GPT Graph Pieces

The graph below uses standard Scorch functions for the same pieces used in a
small GPT-style block. Embeddings turn character ids and positions into numeric
vectors. Causal self-attention lets each character look only at earlier
characters, which is what makes next-character prediction valid. Residual paths
are shortcut connections that help information move through the model.

```{r blocks}
token_positions <- function(idx, block_size) {
  batch <- idx$shape[1]
  positions <- torch_arange(1, block_size, dtype = torch_long())
  positions$unsqueeze(1)$expand(c(batch, block_size))
}

scores_for_cross_entropy <- function(scores) scores$transpose(2, 3)
```

## Scorcher Graph

```{r model}
vocab_size <- length(chars)
n_embd <- 192L
n_head <- 6L

model <- initiate_scorch(dl) |>
  scorch_input(tokens) |>
  scorch_layer(embedding,
               num_embeddings = vocab_size,
               embedding_dim = n_embd,
               .from = tokens,
               .name = token_emb) |>
  scorch_function(token_positions,
                  block_size = block_size,
                  .from = tokens,
                  .name = positions) |>
  scorch_layer(embedding,
               num_embeddings = block_size,
               embedding_dim = n_embd,
               .from = positions,
               .name = position_emb) |>
  scorch_add_skip(.from = c(token_emb, position_emb), .name = embeddings) |>
  scorch_dropout(p = 0.1, .from = embeddings, .name = embedding_drop) |>
  scorch_block(
    layer_norm(normalized_shape = n_embd),
    multihead_attention(
      embed_dim = n_embd,
      num_heads = n_head,
      dropout = 0.1,
      batch_first = TRUE,
      causal = TRUE,
      self_attention = TRUE
    ),
    dropout(p = 0.1),
    residual = TRUE,
    .from = embedding_drop,
    .name = attention_block
  ) |>
  scorch_block(
    layer_norm(normalized_shape = n_embd),
    linear(in_features = n_embd, out_features = 4 * n_embd),
    gelu(),
    linear(in_features = 4 * n_embd, out_features = n_embd),
    dropout(p = 0.1),
    residual = TRUE,
    .from = attention_block,
    .name = mlp_block
  ) |>
  scorch_layer(layer_norm, normalized_shape = n_embd,
               .from = mlp_block, .name = final_norm) |>
  scorch_layer(linear,
               in_features = n_embd,
               out_features = vocab_size,
               .from = final_norm,
               .name = character_scores) |>
  scorch_function(scores_for_cross_entropy,
                  .from = character_scores,
                  .name = scores_for_loss) |>
  scorch_output(scores_for_loss)

scorch_spec(model)
plot(model, input_shapes = list(tokens = "batch x block_size"))
```

## Training

Character-level GPT training is more computationally intensive than the small
tabular examples. The workflow is the same: compile the graph with
cross-entropy loss, fit it on batches of character ids, and snapshot the run
for auditing. Cross-entropy loss compares the model's raw character scores with
the correct next character.

```{r train}
scorch_fit <- model |>
  compile_scorch(
    loss_fn = nn_cross_entropy_loss(),
    optimizer_fn = optim_adamw,
    optimizer_params = list(lr = 5e-4)
  ) |>
  fit_scorch(num_epochs = 20, seed = 12345, verbose = TRUE)

run <- scorch_snapshot(scorch_fit, data = text)
scorch_audit(run)
```

## Evaluation

First, score the trained model on text it did not train on. Lower validation
loss and perplexity are better. Perplexity is a language-model score where
smaller values mean the model is less surprised by the validation text.
Next-character accuracy is a simple check of how often the model's most likely
next character matches the real next character.

```{r evaluate}
loss_fn <- nn_cross_entropy_loss()
total_loss <- 0
total_tokens <- 0
total_correct <- 0

coro::loop(for (batch in validation_dl) {
  scores <- scorch_predict(scorch_fit, batch$input)
  targets <- batch$output[[1]]

  loss <- loss_fn(scores, targets)
  n_tokens <- prod(as.numeric(targets$shape))

  predictions <- torch_argmax(scores, dim = 2)
  correct <- torch_sum(
    (predictions == targets)$to(dtype = torch_float())
  )$item()

  total_loss <- total_loss + loss$item() * n_tokens
  total_tokens <- total_tokens + n_tokens
  total_correct <- total_correct + correct
})

validation_loss <- total_loss / total_tokens

data.frame(
  validation_loss = validation_loss,
  perplexity = exp(validation_loss),
  next_character_accuracy = total_correct / total_tokens
)
```

Then inspect the model qualitatively by asking it to continue a validation
passage that was not used for training.

```{r seed-text}
seed_tokens <- validation_xy$x[1, ]$unsqueeze(1)
seed_ids <- as.integer(as_array(seed_tokens$squeeze(1)))
seed_text <- paste(chars[seed_ids], collapse = "")

cat(seed_text)
```

```{r generate}
generated <- seed_tokens$clone()
n_new <- 600
temperature <- 0.9

for (i in seq_len(n_new)) {
  current_length <- as.integer(generated$shape[2])
  start <- max(1L, current_length - block_size + 1L)
  context <- generated[, start:current_length]

  scores <- scorch_predict(
    scorch_fit,
    input = list(tokens = context)
  )

  next_scores <- scores[, , as.integer(scores$shape[3])] / temperature
  probabilities <- nnf_softmax(next_scores, dim = 2)
  next_id <- torch_multinomial(probabilities, num_samples = 1)

  generated <- torch_cat(
    list(generated, next_id$to(dtype = torch_long())),
    dim = 2
  )
}

generated_ids <- as.integer(as_array(generated$squeeze(1)))
generated_text <- paste(chars[generated_ids], collapse = "")

cat(generated_text)
```

After a short CPU run, the output may only resemble Shakespeare locally through
common character patterns and short words. Longer training, a larger model, or
GPU training should make the samples more coherent.

The same graph can be converted for use with `luz`. The validation dataloader,
which contains text that was not used for training, can then be passed to
`luz::evaluate()`:

```{r luz}
torch_module <- as_torch(model)
luz_dl <- torch::dataloader(
  torch::tensor_dataset(xy$x, xy$y),
  batch_size = 64,
  shuffle = TRUE
)
luz_validation_dl <- torch::dataloader(
  torch::tensor_dataset(validation_xy$x, validation_xy$y),
  batch_size = 64,
  shuffle = FALSE
)

luz_fit <- torch_module |>
  luz::setup(
    loss = nn_cross_entropy_loss(),
    optimizer = optim_adamw
  ) |>
  luz::fit(data = luz_dl, epochs = 20)

luz::evaluate(luz_fit, data = luz_validation_dl)
```
