---
title: "Fitting Neural Networks with Scorcher using the MNIST Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting Neural Networks with Scorcher using the MNIST Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  error = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette builds a small convolutional neural network for handwritten digit
classification. A convolutional neural network is designed for image data
because it looks for local visual patterns such as edges and curves. MNIST
images are 28 by 28 grayscale pictures of digits from 0 to 9, and the goal is
to assign each image to the correct digit class.

The example focuses on the pieces that are common in image classification:
loading image tensors, adding convolution and pooling layers, flattening the
image representation, training by comparing predicted digit scores with the
true digit labels, and checking both accuracy and example predictions.

<br>

## Setup

### Installing and Loading Required Packages

Before we begin, ensure that you have the necessary packages installed. 
You can install them using the following commands:

```{r install, eval = FALSE}
install.packages("torch")
install.packages("torchvision")
install.packages("scorcher")
```

Additionally, you'll need to install the system dependencies used by R `torch`.
Follow the R `torch` installation instructions
[here](https://torch.mlverse.org/start/installation/). Then load the packages
used in this analysis:

```{r setup}
library(torch)
library(torchvision)
library(scorcher)
```

### Loading the Data

The training images are converted to floating-point tensors and given an
explicit channel dimension. A tensor is the data format used by `torch`; here,
each tensor stores many 28 by 28 images. The labels are stored as integer digit
values, so they use `torch_long()`.

```{r data}
# Training data

train_data <- mnist_dataset(
  root = tempdir(),
  download = TRUE,
  transform = transform_to_tensor
)

x_train <- torch_tensor(train_data$data, dtype = torch_float()) |> 
  torch_unsqueeze(2)

y_train <- torch_tensor(train_data$targets, dtype = torch_long())

```

### Visualizing the Data

Before modeling, it is useful to inspect a few images to confirm that the data
were loaded and reshaped as expected.

```{r plt1}
# Visualizing the MNIST training data

par(mfrow = c(2, 2), mar = c(1, 1, 1, 1), pty = "s")

for (i in 1:4) {
  img <- x_train[i, , drop = F] |> as.array() |> matrix(nrow = 28)
  image(t(img[nrow(img):1, ]), axes = F, col = gray.colors(256))
}
```

## Using Scorcher

### Defining the Neural Network

The graph below follows a standard image classification shape: convolutional
layers look for local visual patterns, pooling makes the image representation
smaller, dropout randomly hides some intermediate values during training, and
fully connected layers return one score per digit.

```{r}
# Create the dataloader

dl <- scorch_create_dataloader(x_train, y_train, batch_size = 500)

# The dataloader feeds the model 500 images at a time.

# Define the neural network

scorch_model <- initiate_scorch(dl) |>
  scorch_input(x) |>
  scorch_layer(conv2d, in_channels = 1, out_channels = 32, kernel_size = 3) |>
  scorch_layer(relu) |>
  scorch_layer(conv2d, in_channels = 32, out_channels = 64, kernel_size = 3) |>
  scorch_layer(relu) |>
  scorch_layer(max_pool2d, kernel_size = 2) |>
  scorch_dropout(p = 0.25) |>
  scorch_flatten() |>
  scorch_layer(linear, in_features = 9216, out_features = 128) |>
  scorch_layer(relu) |>
  scorch_layer(linear, in_features = 128, out_features = 10) |>
  scorch_output() |>
  compile_scorch(
    loss_fn          = nn_cross_entropy_loss(),
    optimizer_fn     = optim_adam,
    optimizer_params = list(lr = 0.001)
  )
```

### Training the Neural Network

The compiled graph is trained directly with `fit_scorch()`. Cross-entropy loss
compares the model's raw digit scores with the correct digit labels.

```{r}
scorch_model <- scorch_model |>
  fit_scorch(num_epochs = 10, verbose = TRUE)
```

The same graph can be converted for use with `luz`:

```{r luz, eval=FALSE}
torch_module <- as_torch(scorch_model)
luz_dl <- torch::dataloader(
  torch::tensor_dataset(x_train, y_train),
  batch_size = 500,
  shuffle = TRUE
)

luz_fit <- torch_module |>
  luz::setup(
    loss = nn_cross_entropy_loss(),
    optimizer = optim_adam
  ) |>
  luz::fit(data = luz_dl, epochs = 10)
```

### Evaluating the Model

Evaluation uses the official MNIST test split. `scorch_evaluate()` handles
evaluation mode, prediction, and accuracy calculation.

```{r}
# Testing data

test_data <- mnist_dataset(
  root = tempdir(),
  train = FALSE,
  transform = transform_to_tensor
)

x_test <- torch_tensor(test_data$data, dtype = torch_float()) |> 
  torch_unsqueeze(2)

y_test <- torch_tensor(test_data$targets, dtype = torch_long())

scorch_evaluate(
  scorch_model,
  input = x_test,
  truth = y_test,
  metric = "accuracy"
)
```

For a model trained with `luz`, use `luz::evaluate()` on a regular torch
dataloader:

```{r luz-evaluate, eval=FALSE}
luz_test_dl <- torch::dataloader(
  torch::tensor_dataset(x_test, y_test),
  batch_size = 500
)

luz::evaluate(luz_fit, data = luz_test_dl)
```

### Visualizing the Predictions 

A few individual predictions make the numeric accuracy easier to interpret.
The title above each image shows the digit predicted by the fitted model.

```{r}
# Visualizing some of the predictions

par(mfrow = c(2, 2), mar = c(1, 1, 1.5, 1), pty = "s")

predictions <- scorch_predict(scorch_model, x_test[1:4, , , ])
predicted_classes <- torch_argmax(predictions, dim = 2) - 1

for (i in 1:4) {
  img <- x_test[i, , drop = F] |> as.array()
  img_mat <- matrix(img, nrow = 28)
  image(
    t(img_mat[nrow(img_mat):1, ]),
    axes = FALSE,
    col = gray.colors(256),
    main = paste("Predicted Value:", predicted_classes[i]$item())
  )
}
```

## Conclusion

This example showed how to build an image classifier with Scorcher: load image
data, define a convolutional graph, train by comparing digit scores with known
labels, evaluate accuracy, and inspect individual predictions. The same
structure can be adapted to other small image classification datasets by
changing the input shape and number of output classes.

## References

**MNIST dataset:**

+ Deng, L. (2012). The mnist database of handwritten digit images for machine learning research. IEEE Signal Processing Magazine, 29(6), 141–142.
