---
title: "Adding layers with scorch_layer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding layers with scorch_layer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(scorcher)
```

## Introduction

This vignette introduces the basic graph-building workflow in `scorcher`.
The central idea is to describe the model as a sequence of named nodes: inputs
enter the graph, layers and functions transform those inputs, and one or more
outputs define what the model returns during training or prediction.

In this vignette, a graph means a diagram-like recipe for the model, and a node
means one named step in that recipe. Naming the steps makes the model easier to
plot, inspect, and connect.

The examples use the small `mtcars` dataset so that the full workflow is easy
to inspect: prepare numeric data for `torch`, define the graph, plot the
architecture, train the model, and compare predictions with observed values.

## Prerequisites

Load `torch` for numeric model data and neural-network layers, `scorcher` for
graph construction and training, and `tidyverse` for the small amount of data
preparation and plotting used here.

```{r}
library(scorcher)
library(torch)
library(tidyverse)
```

## Naming Nodes

In scorcher's graph-based architecture, every node needs a unique name.
These names are how the graph connects steps together. When you write
`.from = c(act2, act1)`, the graph knows exactly which upstream
outputs to pass forward. This is what enables multi-input models, parallel
branches, skip connections, and fusion architectures.

Names are arbitrary -- you can use whatever makes sense to you. The only
rule is that each name must be unique within the model. That said,
descriptive names help with debugging because they appear in error messages
and in `plot()` output.

Here are some common naming conventions:

| Prefix | Meaning | Layers |
|--------|---------|--------|
| `fc`   | fully connected | `nn_linear` |
| `conv` | convolution, used mostly for image or sequence patterns | `nn_conv1d`, `nn_conv2d` |
| `act`  | nonlinear step, such as ReLU or GELU | `nn_relu`, `nn_gelu`, `nn_sigmoid` |
| `pool` | pooling, used to shrink image or sequence representations | `nn_max_pool2d`, `nn_avg_pool1d` |
| `bn`   | batch normalization, used to stabilize intermediate values | `nn_batch_norm1d` |
| `drop` | dropout, which randomly hides some values during training | `nn_dropout` |
| `flat` | flatten | `nn_flatten` |
| `attn` | attention | `nn_multihead_attention` |
| `enc`  | encoder | `nn_transformer_encoder_layer` |
| `res`  | residual/skip | `scorch_add_skip`, `scorch_block` |
| `emb`  | embedding, used to turn ids into numeric vectors | `nn_embedding` |

Multi-head attention can be added with `scorch_layer()` directly. Attention is
a layer that lets each position in a sequence compare itself with other
positions. Pass the query, key, and value nodes through `.from`.

```{r attention-example, eval=FALSE}
attention_model <- initiate_scorch() |>
  scorch_input(embeddings) |>
  scorch_layer(multihead_attention,
               embed_dim = 64,
               num_heads = 4,
               .from = c(embeddings, embeddings, embeddings),
               .name = self_attention) |>
  scorch_output(self_attention)
```

For repeated pieces, use `scorch_block()` with unquoted layer calls. This keeps
common architecture patterns readable while still expanding to ordinary graph
nodes.

```{r block-example, eval=FALSE}
model <- initiate_scorch() |>
  scorch_input(features) |>
  scorch_layer(linear, in_features = 10, out_features = 64, .name = hidden) |>
  scorch_block(
    linear(in_features = 64, out_features = 64),
    gelu(),
    residual = TRUE,
    repeats = 2,
    .from = hidden,
    .name = residual_stack
  ) |>
  scorch_output(residual_stack)
```

Add a number suffix when you have multiples of the same type: `conv1`,
`conv2`, `fc1`, `fc2`, and so on.

## Data Preparation: `mtcars` Dataset

For this vignette, we use the `mtcars` dataset. The response is miles per
gallon (`mpg`), and the predictors are horsepower (`hp`) and weight (`wt`).
The predictors are standardized before training so that both inputs are on a
similar scale.

```{r}
# Prepare the data

data(mtcars)

dat <- mtcars |>
  
  select(mpg, hp, wt) |>
  
  mutate(
    
    hp = scale(hp), 
    
    wt = scale(wt))

# Convert to torch tensors, the numeric data format used by torch

x <- dat[, c("hp", "wt")] |>
  
  as.matrix() |> 
  
  torch_tensor(dtype = torch_float())

y <- torch_tensor(mtcars$mpg, dtype = torch_float())

y <- y$unsqueeze(2)
```

## Building a Simple Neural Network

The model begins with one input node named `x`. Each `scorch_layer()` call adds
the next model step. Because these layers form a straight sequence, only the
input and output need explicit names.

```{r}

# Create the dataloader

dl <- scorch_create_dataloader(x, y, batch_size = 32)

# Initialize the Scorcher model

scorch_model <- initiate_scorch(dl) |>
  
  scorch_input(x) |>
  
  scorch_layer(linear, in_features = 2, out_features = 16) |>
  
  scorch_layer(relu) |>
  
  scorch_layer(linear, in_features = 16, out_features = 1) |>
  
  scorch_output()

# View the model architecture

print(scorch_model)
```

This network consists of:

- A hidden layer with 16 units and ReLU, a common nonlinear step.
- An output layer with 1 unit (to predict `mpg`).

## Adding a Residual Block

A residual connection adds the input to a block back to the block's output.
This gives the model a shortcut path and can make deeper models easier to
train. In Scorcher, `scorch_block()` is a compact way to write that pattern.

```{r}

# Create and view updated model

scorch_model <- initiate_scorch(dl) |>
  
  scorch_input(x) |>
  
  scorch_layer(linear, in_features = 2, out_features = 16, .name = hidden) |>
  
  scorch_block(
    linear(in_features = 16, out_features = 16),
    relu(),
    residual = TRUE,
    .from = hidden,
    .name = residual
  ) |>
  
  scorch_layer(linear, in_features = 16, out_features = 1) |>
  
  scorch_output()

print(scorch_model)
  
```

This model includes:

- Two linear layers with ReLU nonlinear steps.
- A residual connection that adds the input directly to the output of the block.

## Training the Model

Training starts by attaching a loss function and optimizer to the graph. The
loss is the number the model tries to make smaller. For a continuous outcome
such as `mpg`, mean squared error measures the average squared prediction
error. The optimizer is the rule used to update model weights; Adam is a useful
default.

```{r}

# Compile and fit the neural network

scorch_model <- scorch_model |>
  
  compile_scorch(
    loss_fn          = nn_mse_loss(),
    optimizer_fn     = optim_adam,
    optimizer_params = list(lr = 0.001)
  ) |>
  
  fit_scorch(num_epochs = 300, verbose = FALSE)

# Or convert the same graph for use with luz:
torch_module <- as_torch(scorch_model)
luz_dl <- torch::dataloader(
  torch::tensor_dataset(x, y),
  batch_size = 32,
  shuffle = TRUE
)

luz_fit <- torch_module |>
  luz::setup(
    loss = nn_mse_loss(),
    optimizer = optim_adam
  ) |>
  luz::fit(data = luz_dl, epochs = 300)

```

## Evaluating the Model

After training, evaluate the fitted model by comparing predicted `mpg` values
with the observed values. `scorch_evaluate()` handles evaluation mode and
prediction internally, so the model object can be used directly.

```{r}
# Make predictions

scorch_evaluate(
  scorch_model,
  input = x,
  truth = y,
  metric = "mse"
)

y_pred <- scorch_predict(scorch_model, x) |> as_array() |> as.numeric()
```

For a model trained with `luz`, use `luz::evaluate()` on a regular torch
dataloader. A dataloader feeds the model small groups of examples at a time:

```{r luz-evaluate, eval=FALSE}
luz::evaluate(luz_fit, data = luz_dl)
```

```{r}
# Plot the actual vs predicted mpg

plt_dat <- data.frame(actual = mtcars$mpg, predicted = y_pred)

ggplot(plt_dat, aes(x = actual, y = predicted)) +
  
  theme_minimal() +
  
  geom_point() +
  
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "black") +
  
  labs(
  
    title = "Actual vs Predicted MPG", 
    
    x = "Actual MPG", 
    
    y = "Predicted MPG")
```

The plot shows how well the model's predictions align with the actual `mpg` values. A perfect model would have all points lying on the dashed line.

## Conclusion

This vignette covered the core `scorcher` workflow: prepare numeric data for
`torch`, declare a model graph, inspect the architecture, train the model, and
evaluate predictions. The same pattern extends from simple regression models to
branched models, residual blocks, attention layers, and other graph structures.
